The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-11-18 sections/intro.tex
Final comments from Arjen

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.

>>>>

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}. The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}). We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}. Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle. To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.

### 2013-11-18 sections/model.tex
Final comments from Arjen

A pseudo-plastic sliding law \citep[supplement]{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a normal distribution along a central (input) value. The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inpress}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

>>>>

A pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a normal distribution along a central (input) value. The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-2013}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
